---
title: "Age cohort calculations"
params:
  prepared_by:
    label: "Prepared by:"
    value: "Richard Martin-Nielsen"
    input: text
date: "`r Sys.Date()`"
output: distill::distill_article
base_url: https://richardmn.github.io/lt_covid_calcs
twitter:
  site: "@rmartinnielsen"
creative_commons: CC BY
---

## Data source

These charts are drawn using data published by the [Official Statistics
Portal](https://osp.stat.gov.lt/pradinis) (OSP) on their [COVID-19 open
data](https://experience.arcgis.com/experience/cab84dcfe0464c2a8050a78f817924ca/page/page_5/)
site, along with the [annual population counts for Lithuanian
municipalities](https://osp.stat.gov.lt/en_GB/gyventojai1), also
published by the OSP.

The R markdown source is available as a [github
repo](https://github.com/RichardMN/lt_covid_calcs).

```{r setup, include=FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(lubridate)
library(zoo)
library(RColorBrewer)
library(httr)
library(jsonlite)
library(readxl)

# Set-up output ----
# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
#knitr::opts_knit$set(base.dir = stringr::str_c(here::here(), "/docs/"), base.url = "/lt_covid_calcs/") # project root folder
knitr::opts_chunk$set(
#	fig.path = stringr::str_c(paste0(paste("images", sep = "/")), "/"),
	message = FALSE,
	warning = FALSE
)
# Load data ----

aggregate_url <-"https://services3.arcgis.com/MF53hRPmwfLccHCj/arcgis/rest/services/Agreguoti_COVID19_atvejai_ir_mirtys/FeatureServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json"

age_sex_pop_url <-     "https://osp.stat.gov.lt/documents/10180/8416503/Resident_population_sex_and_age_2021.xls"

download_restful <- function (api_call = "",
                              data_filter = function(x) { return(x) }) {
  if(str_length(api_call) == 0) {
    message("Empty api_call")
    return(NULL)
  }
  get_data <- GET(api_call)

  get_content <- content(get_data, "text")

  get_json <- fromJSON(get_content, flatten = TRUE)

  get_df <- as.data.frame(get_json$features)

  result_offset <- dim(get_df)
  result_offset <- result_offset[1]

  get_df <- data_filter(get_df)

  page <- 1
  done_download <- FALSE
  supp_data <- get_data

  rm("get_data", "get_content", "get_json")

  while (!done_download && supp_data$status_code == 200) {
    offset <- result_offset * page
    page <- page + 1
    supp_api_call <-
      paste(api_call, "&resultOffset=", sprintf("%d", offset), sep = "")
    supp_data <- GET(supp_api_call)
    message( ".", appendLF = FALSE)

    supp_content <- content(supp_data, "text")
    supp_json <- fromJSON(supp_content, flatten = TRUE)
    supp_df <- as.data.frame(supp_json$features) %>%
      data_filter()
    row_count <- dim(supp_json$features)
    row_count <- row_count[1]

    if (!hasName(supp_json, "exceededTransferLimit")) {
      message(
        "\nDownload complete"
      )
      done_download <- TRUE
    }
    if (supp_data$status_code == 200) {
      get_df <- rbind(get_df, supp_df)
    } else {
      message("\nDownload finished with unexpected status code",
              supp_data$status_code
      )
      break
    }
  }
  return(get_df)
}

#' Download Excel Documents
#' Taken from covidregionaldata library
#'
#' @param url Character string containing the full URL to the Excel document.
#' @param archive Character string naming the file name to assign in the
#' temporary directory.
#' @param transpose Logical, should the read in data be transposed
#' @param ... Additional parameters to pass to `read_excel()`.
#' @inheritParams message_verbose
#' @importFrom readxl read_excel
#' @return A `data.frame`.
download_excel <- function(url, archive, verbose = FALSE,
                           transpose = TRUE, ...) {
  # download
  archive <- file.path(tempdir(), archive)
  download.file(
    url = url,
    destfile = archive,
    mode = "wb", quiet = !(verbose)
  )
  # read in
  dt <- suppressMessages(
    read_excel(archive, ...)
  )

  if (transpose) {
    dt <- t(dt)
  }
  dt <- as.data.frame(dt)
  return(dt)
}

lt_aggregate <- download_restful(
  api_call = aggregate_url,
  data_filter = function(x){
    transmute(x,
              municipality_name = attributes.municipality_name,
              date = attributes.date,
              sex = attributes.sex,
              age_gr = attributes.age_gr,
              new_cases = attributes.new_cases,
              deaths_all = attributes.deaths_all,
              deaths_cov1 = attributes.deaths_cov1,
              deaths_cov2 = attributes.deaths_cov2,
              deaths_cov3 = attributes.deaths_cov3
    )})

  lt_age_sex_data <- download_excel(
    age_sex_pop_url,
    archive = "population_data",
    transpose = FALSE,
    sheet = "Total",
    col_types = c("text",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric"),
    skip = 7) %>%
    rename(location = "...1", total = "...2")


```

```{r data_manipulation, echo=TRUE}
age_bands_municipalities <- tibble(lt_aggregate) %>%
#  select(-object_id) %>%
  mutate(date = as_date(date)) %>%
  group_by(municipality_name, date, age_gr) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE))) %>%
  ungroup()

age_bands <- age_bands_municipalities %>%
  group_by(date, age_gr) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)))

natl_age_data <-  lt_age_sex_data %>%
  filter( location == "Total" ) %>%
  select(-location) %>%
  pivot_longer(
    cols = !"total",
    values_to = "count",
    names_to = "age_range") %>%
  mutate_if(is.character, str_remove_all, pattern = "\\d+[^\\d]") %>%
  mutate(age_range = as.numeric(age_range)) %>%
  mutate(age_range = replace_na(age_range, 85)) %>%
  mutate(cohort = cut(age_range,
                      c(0, 9, 19, 29, 39, 49, 59, 69, 79, 89 ),
                      c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+"),
                      include.lowest = TRUE)) %>%
  select(-age_range,-total) %>%
  group_by(cohort) %>%
  summarise(count = sum(count))

lt_age_data <- lt_age_sex_data %>%
  filter(grepl("mun.$", location)) %>%
  pivot_longer(
    cols = !c("location","total"),
    values_to = "count",
    names_to = "age range") %>%
  mutate_if(is.character, str_replace_all, pattern = " mun.", replacement = "")
age_band_factors <- age_bands %>%
  mutate(cohort = case_when(
    age_gr == "80-89" ~ "80+",
    age_gr == "90-99" ~ "80+",
    age_gr == "Centenarianai" ~ "80+",
    age_gr == "Nenustatyta" ~ NA_character_,
    TRUE ~ age_gr
  )) %>%
  filter(!is.na(cohort)) %>%
  select(-age_gr) %>%
  group_by(date, cohort) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE))) %>%
  ungroup()

per_capita_rates <- left_join(age_band_factors, natl_age_data, by = c("cohort")) %>%
  mutate(population = count,
         cases_per_100k = new_cases / count * 100000,
         deaths_all_per_mill = deaths_all / count * 1000000) %>%
  select(-count)

```


```{r age_band_graphs, echo=FALSE}

colourCount = length(unique(age_bands$age_gr))
getPalette = colorRampPalette(brewer.pal(8, "Set2"))

age_bands %>%
  group_by(age_gr) %>%
  mutate(cases_7d = zoo::rollmean(new_cases,k=7, fill=0) ) %>%
  ungroup() %>%
  filter(date > ymd("2021-08-01")) %>%
  ggplot(aes(x = date, y=cases_7d, fill=age_gr)) +
  theme_minimal() +
  geom_col(width=1, position = position_stack(reverse = TRUE)) +
  #scale_fill_brewer(palette = "Set2") +
  scale_fill_viridis_d(
    name = "Age cohort",
    option = "inferno",
    breaks = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69",
               "70-79", "80-89", "90-99", "Centenarianai",  "Nenustatyta"),
    labels = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69",
               "70-79", "80-89", "90-99", "100+",  "Unknown")
  ) +
  scale_y_continuous(sec.axis = dup_axis()) +
  labs(title="Lithuania - COVID-19 cases by age group",
       subtitle="7 day rolling average",
       y="New cases",
       x="Date",
       caption="Richard Martin-Nielsen | Data: Office of Statistics Portal osp.stat.gov.lt") +
  scale_x_date()
```

These charts are inspired by the narrow age cohort graph given by the OSP
on their [pandemic illustrations
page](https://osp.stat.gov.lt/pandemijos-iliustracijos):

`r paste0("![Figure from https://osp.stat.gov.lt](https://osp.stat.gov.lt/documents/10180/8555211/vaiku_atvejai_per14d_100k_", strftime(today(), "%Y%m%d"),".png){.external}")`

Because the age cohorts given in the two sources used do not align,
when calculating COVID rates relative to age cohorts, a smaller
number of larger cohorts is used. It is also possible to extend the graph
further back into 2021.

```{r cohort_prevalance_cases_cumulative, echo=FALSE, preview=TRUE}
per_capita_rates %>%
  group_by(cohort) %>%
  mutate(pc_cases_14d = zoo::rollsum(cases_per_100k,k=14, fill=0, align="right") ) %>%
  ungroup() %>%
  filter(date > ymd("2020-11-01")) %>%
  ggplot(aes(x = date, y=pc_cases_14d, colour=cohort)) +
  theme_minimal() +
  geom_line(size=1) +
  #scale_fill_brewer(palette = "Set2") +
  scale_colour_viridis_d(name = "Age cohort", option = "inferno") +
  scale_y_continuous(sec.axis = sec_axis(~ .)) +
  labs(title="Lithuania - COVID-19 cases by age group",
       subtitle="14 day cumulative per 100,000",
       y="New cases",
       x="Date",
       caption="Richard Martin-Nielsen | Data: Office of Statistics Portal osp.stat.gov.lt") +
  scale_x_date()
```

```{r cohort_prevalance_cases_mean, echo=FALSE}
per_capita_rates %>%
  group_by(cohort) %>%
  mutate(pc_cases_7d_mean = zoo::rollmean(cases_per_100k,k=7, fill=0, align="right") ) %>%
  ungroup() %>%
  filter(date > ymd("2020-11-01")) %>%
  ggplot(aes(x = date, y=pc_cases_7d_mean, colour=cohort)) +
  theme_minimal() +
  geom_line(size=1) +
  #scale_fill_brewer(palette = "Set2") +
  scale_colour_viridis_d(name = "Age cohort", option = "inferno") +
  scale_y_continuous(sec.axis = sec_axis(~ .)) +
  labs(title="Lithuania - COVID-19 cases by age group",
       subtitle="7 day rolling average per 100,000",
       y="New cases",
       x="Date",
       caption="Richard Martin-Nielsen | Data: Office of Statistics Portal osp.stat.gov.lt") +
  scale_x_date()
```

```{r cohort_prevalance_deaths_mean, echo=FALSE}
per_capita_rates %>%
  group_by(cohort) %>%
  mutate(pc_deaths_7d_mean = zoo::rollmean(deaths_all_per_mill,k=7, fill=0, align="right") ) %>%
  ungroup() %>%
  filter(date > ymd("2020-11-01")) %>%
  ggplot(aes(x = date, y=pc_deaths_7d_mean, colour=cohort)) +
  theme_minimal() +
  geom_line(size=1) +
  #scale_fill_brewer(palette = "Set2") +
  scale_colour_viridis_d(name = "Age cohort", option = "inferno") +
  scale_y_continuous(sec.axis = sec_axis(~ .)) +
  labs(title="Lithuania - COVID-19 deaths by age group",
       subtitle="7 day average per million",
       y="Deaths",
       x="Date",
       caption="Richard Martin-Nielsen | Data: Office of Statistics Portal osp.stat.gov.lt") +
  scale_x_date()

```
