---
title: "Hospitalisations"
params:
  prepared_by:
    label: "Prepared by:"
    value: "Richard Martin-Nielsen"
    input: text
date: "`r Sys.Date()`"
output: distill::distill_article
base_url: https://richardmn.github.io/lt_covid_calcs
twitter:
  site: "@rmartinnielsen"
creative_commons: CC BY
---

## Data source

These charts are drawn using data published by the [Official Statistics
Portal](https://osp.stat.gov.lt/pradinis) (OSP) on their [COVID-19 open
data](https://experience.arcgis.com/experience/cab84dcfe0464c2a8050a78f817924ca/page/page_5/)
site, along with the [annual population counts for Lithuanian
municipalities](https://osp.stat.gov.lt/en_GB/gyventojai1), also
published by the OSP.

The R markdown source is available as a [github
repo](https://github.com/RichardMN/lt_covid_calcs).

```{r setup, include=FALSE}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(lubridate)
library(zoo)
library(RColorBrewer)
library(httr)
library(jsonlite)
library(readxl)
library(ggrepel)

# Set-up output ----
# Figure path on disk = base.dir + fig.path
# Figure URL online = base.url + fig.path
#knitr::opts_knit$set(base.dir = stringr::str_c(here::here(), "/docs/"), base.url = "/lt_covid_calcs/") # project root folder
knitr::opts_chunk$set(
#	fig.path = stringr::str_c(paste0(paste("images", sep = "/")), "/"),
	message = FALSE,
	warning = FALSE
)
# Load data ----

aggregate_url <-"https://services3.arcgis.com/MF53hRPmwfLccHCj/arcgis/rest/services/Agreguoti_COVID19_atvejai_ir_mirtys/FeatureServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=json"

age_sex_pop_url <-     "https://osp.stat.gov.lt/documents/10180/8416503/Resident_population_sex_and_age_2021.xls"

hosp_data_url <- "https://opendata.arcgis.com/api/v3/datasets/f8f5c84b20e24ff99be4dd472d377640_0/downloads/data?format=csv&spatialRefId=4326"

download_restful <- function (api_call = "",
                              data_filter = function(x) { return(x) }) {
  if(str_length(api_call) == 0) {
    message("Empty api_call")
    return(NULL)
  }
  get_data <- GET(api_call)

  get_content <- content(get_data, "text")

  get_json <- fromJSON(get_content, flatten = TRUE)

  get_df <- as.data.frame(get_json$features)

  result_offset <- dim(get_df)
  result_offset <- result_offset[1]

  get_df <- data_filter(get_df)

  page <- 1
  done_download <- FALSE
  supp_data <- get_data

  rm("get_data", "get_content", "get_json")

  while (!done_download && supp_data$status_code == 200) {
    offset <- result_offset * page
    page <- page + 1
    supp_api_call <-
      paste(api_call, "&resultOffset=", sprintf("%d", offset), sep = "")
    supp_data <- GET(supp_api_call)
    message( ".", appendLF = FALSE)

    supp_content <- content(supp_data, "text")
    supp_json <- fromJSON(supp_content, flatten = TRUE)
    supp_df <- as.data.frame(supp_json$features) %>%
      data_filter()
    row_count <- dim(supp_json$features)
    row_count <- row_count[1]

    if (!hasName(supp_json, "exceededTransferLimit")) {
      message(
        "\nDownload complete"
      )
      done_download <- TRUE
    }
    if (supp_data$status_code == 200) {
      get_df <- rbind(get_df, supp_df)
    } else {
      message("\nDownload finished with unexpected status code",
              supp_data$status_code
      )
      break
    }
  }
  return(get_df)
}

#' Download Excel Documents
#' Taken from covidregionaldata library
#'
#' @param url Character string containing the full URL to the Excel document.
#' @param archive Character string naming the file name to assign in the
#' temporary directory.
#' @param transpose Logical, should the read in data be transposed
#' @param ... Additional parameters to pass to `read_excel()`.
#' @inheritParams message_verbose
#' @importFrom readxl read_excel
#' @return A `data.frame`.
download_excel <- function(url, archive, verbose = FALSE,
                           transpose = TRUE, ...) {
  # download
  archive <- file.path(tempdir(), archive)
  download.file(
    url = url,
    destfile = archive,
    mode = "wb", quiet = !(verbose)
  )
  # read in
  dt <- suppressMessages(
    read_excel(archive, ...)
  )

  if (transpose) {
    dt <- t(dt)
  }
  dt <- as.data.frame(dt)
  return(dt)
}

lt_aggregate <- download_restful(
  api_call = aggregate_url,
  data_filter = function(x){
    transmute(x,
              municipality_name = attributes.municipality_name,
              date = attributes.date,
              sex = attributes.sex,
              age_gr = attributes.age_gr,
              new_cases = attributes.new_cases,
              deaths_all = attributes.deaths_all,
              deaths_cov1 = attributes.deaths_cov1,
              deaths_cov2 = attributes.deaths_cov2,
              deaths_cov3 = attributes.deaths_cov3
    )})

  lt_age_sex_data <- download_excel(
    age_sex_pop_url,
    archive = "population_data",
    transpose = FALSE,
    sheet = "Total",
    col_types = c("text",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric", "numeric", "numeric", "numeric",
                  "numeric"),
    skip = 7) %>%
    rename(location = "...1", total = "...2")

  
  #https://open-data-ls-osp-sdg.hub.arcgis.com/datasets/f8f5c84b20e24ff99be4dd472d377640_0/about
  
lt_hosp_data_raw <- vroom::vroom(hosp_data_url)

caption_text <- paste0(
         "Richard Martin-Nielsen | Data: Office of Statistics Portal osp.stat.gov.lt\nPrepared ",
         Sys.Date())

```

```{r data_manipulation, echo=TRUE}
age_bands_municipalities <- tibble(lt_aggregate) %>%
#  select(-object_id) %>%
  mutate(date = as_date(date)) %>%
  group_by(municipality_name, date, age_gr) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)), .groups = "drop")

age_bands <- age_bands_municipalities %>%
  group_by(date, age_gr) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)), .groups = "drop")

natl_age_data <-  lt_age_sex_data %>%
  filter( location == "Total" ) %>%
  select(-location) %>%
  pivot_longer(
    cols = !"total",
    values_to = "count",
    names_to = "age_range") %>%
  mutate_if(is.character, str_remove_all, pattern = "\\d+[^\\d]") %>%
  mutate(age_range = as.numeric(age_range)) %>%
  mutate(age_range = replace_na(age_range, 85)) %>%
  mutate(cohort = cut(age_range,
                      c(0, 9, 19, 29, 39, 49, 59, 69, 79, 89 ),
                      c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+"),
                      include.lowest = TRUE)) %>%
  select(-age_range,-total) %>%
  group_by(cohort) %>%
  summarise(count = sum(count), .groups = "drop")

lt_age_data <- lt_age_sex_data %>%
  filter(grepl("mun.$", location)) %>%
  pivot_longer(
    cols = !c("location","total"),
    values_to = "count",
    names_to = "age range") %>%
  mutate_if(is.character, str_replace_all, pattern = " mun.", replacement = "")
age_band_factors <- age_bands %>%
  mutate(cohort = case_when(
    age_gr == "80-89" ~ "80+",
    age_gr == "90-99" ~ "80+",
    age_gr == "Centenarianai" ~ "80+",
    age_gr == "Nenustatyta" ~ NA_character_,
    TRUE ~ age_gr
  )) %>%
  filter(!is.na(cohort)) %>%
  select(-age_gr) %>%
  group_by(date, cohort) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)), .groups = "drop")

per_capita_rates <- left_join(age_band_factors, natl_age_data, by = c("cohort")) %>%
  mutate(population = count,
         cases_per_100k = new_cases / count * 100000,
         deaths_all_per_mill = deaths_all / count * 1000000) %>%
  select(-count)

lt_hosp_data <- lt_hosp_data_raw %>%
  select(-X,-Y,-object_id) %>%
  group_by(date) %>%
    summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)), .groups = "drop") %>%
  pivot_longer(
    cols = !c("date"),
    values_to = "count",
    names_to = "event")

lt_hosp_data %>%
  filter(event == "cv19_hospital_beds_occupied_r7m") %>%
  ggplot(aes(x=date,y=count)) + geom_line()

lt_delta <- lt_hosp_data %>%
  filter(event == "cv19_hospital_beds_occupied_r7m") %>%
  arrange(date) %>%
  mutate(delta = count - lag(count), 
         event = "cv19_hospital_beds_occupied_r7mean_delta") %>%
  mutate(count=delta) %>%
  select(-delta)

# compare the calculated deltas with the admissions data - they don't line up
# ggplot() + geom_line(data=lt_delta, mapping=aes(x=date,y=count)) + geom_line(lt_hosp_cv19%>%filter(event=="Admissions"), mapping=aes(x=date,y=count))

lt_hosp_data <- bind_rows(lt_hosp_data, lt_delta)  
```

```{r hospitalisation_graphs, echo=FALSE}

lt_hosp_cv19 <- lt_hosp_data %>%
  filter(str_detect(event, "^cv19.*admissions")
         |str_detect(event, "^cv19.*in_use")
         |str_detect(event, "^cv19.*delta")
         |str_detect(event, "^cv19.*occupied")) %>%
  mutate( event = gsub("cv19_","", event)) %>%
  mutate( event = gsub("_r7m.*","", event)) %>%
  mutate( event = str_to_title(gsub("_"," ", event)))
events_to_plot <- c(
              # "Hospital Beds Occupied",
              # "Oxygen Beds Occupied",
              # "Mv Beds Occupied",
              # "Icu Beds Occupied",
              # "Icu Mv Beds Occupied", 
              # "Icu Ecmo In Use",
              "Admissions", 
              "Admissions Oxygen",
              "Admissions Mv"
              )

lt_hosp_cv19 %>%
  filter( event %in% events_to_plot) %>%
  filter(!(event == "Admissions" & count == 0)) %>%
  ggplot(aes(x=date,y=count, colour=event)) +
  geom_line(size=1) +
    geom_text_repel(aes(x=date,y=count,label=event,colour=event),
                  nudge_x=75,
                  direction="y",hjust="left",
                  data = arrange(lt_hosp_cv19,event,date)%>% 
                    filter( event %in% events_to_plot)%>%
                    group_by(event)%>%
                    summarise(across(everything(), last))) +
    theme_minimal() +
    scale_colour_viridis_d() +
  theme(legend.position = "none") +
    labs(title="Lithuania - Hospitalisation data",
       subtitle="7 day means",
       y="Beds",
       x="Date",
       caption=caption_text
         ) +
   scale_x_date(limits = c(ymd("2020-12-01",NA)),expand = expansion(add = c(0,100))) 
```



```{r normalised_hospitalisation, echo=FALSE}
lt_case_deaths <- lt_aggregate %>%
  group_by(date) %>%
  summarise(across(where(is.numeric), ~ sum(.x, na.rm=TRUE)), .groups = "drop") %>%
  mutate(date = as_date(date)) %>%
  pivot_longer(
    cols = !c("date"),
    values_to = "count",
    names_to = "event") %>%
    mutate( event = str_to_title(gsub("_"," ", event))) %>%
  group_by(event) %>%
  mutate(count = zoo::rollmean(count,k=7, fill=NA, align="right") ) %>%
  ungroup()

lt_hosp_cv19_max <- lt_hosp_cv19 %>%
  bind_rows(lt_case_deaths) %>%  
  filter(date > ymd("2020-11-01") & date < ymd("2021-7-01") ) %>%
  group_by(event) %>%
  summarise(max = max(count)) %>%
  # fudge things slightly: admissions must be > admissions O2 so we'll set it to 250
  mutate(max = if_else(event == "Admissions", 250, max))
  
lt_hosp_nrml <- left_join(bind_rows(lt_hosp_cv19, lt_case_deaths), lt_hosp_cv19_max, by=c("event")) %>%
  mutate( count = case_when(
    event=="Admissions" && count == 0  ~ NA_real_,
    TRUE ~ count ) ) %>%
  filter(!(event == "Admissions" & count == 0)) %>%
  mutate(normalised= count / max * 100) %>%
    filter(event %in% c(
      #"Icu Ecmo In Use",
                        "Admissions",
                        "Admissions Oxygen",
                        "Admissions Mv",
                      "New Cases",
                      "Deaths Cov1"
                      #"Icu Ecmo In Use"
                      )) %>%
  #filter(!(event=="Admissions" && normalised == 0)) %>%
  arrange(date)

lt_hosp_nrml %>%
  ggplot(aes(x=date,y=normalised, colour=event)) +
  geom_line(size=1) +
    geom_text_repel(aes(x=date,y=normalised,label=event,colour=event),
                  nudge_x=100,
                  direction="y",hjust="right",
                  data = arrange(lt_hosp_nrml,event,date)%>%group_by(event)%>%summarise(across(everything(), last))) +
                  #data=group_by(lt_hosp_nrml%>%filter(date == ymd("2022-01-13")), event)%>%tail(6)) +
  geom_hline(yintercept = 100) +
  theme_minimal() +
    scale_colour_viridis_d() +
  theme(legend.position = "none") +
    labs(title="Lithuania - Case, Hospitalisation and Deaths data",
       subtitle="7 day means - maximum of December 2020 wave = 100 (Estimate used for Admissions)",
       y="",
       x="Date",
       caption=caption_text) +
  scale_x_date(limits = c(ymd("2020-12-01",NA)),expand = expansion(add = c(0,100))) +
  scale_y_continuous(limits = c(0,NA))
```


## Dataset description (google translation) {.appendix}


*google translation of original Lithuanian*^[https://open-data-ls-osp-sdg.hub.arcgis.com/datasets/f8f5c84b20e24ff99be4dd472d377640_0/about]

7-day averages for available and occupied inpatient beds, oxygen-supplied beds, beds in intensive care and resuscitation wards, artificial lung ventilation equipment and its use, ECMO systems and their occupancy, patient flow, and amounts of Remdesivir.

7-day averages for available and occupied inpatient beds, oxygen-supplied beds, beds in intensive care and resuscitation wards, artificial lung ventilation equipment and its use, ECMO systems and their occupancy, patient flow, and amounts of Remdesivir.

Data is updated in the morning at 10:00.

Description of information provided:
`date `- reporting day (end date of the reporting period 16: 00-16: 00)
`hospital_name `- the name of the hospital
`healthcare_region `- region (five territorial health insurance funds)
`hospital_beds_total_r7mean`- total number of inpatient hospital beds (without RITS!)
`hospital_beds_occupied_r7mean`- number of inpatient hospital beds (without RITS!)
`oxygen_beds_total_r7mean`- all beds with oxygen supply (part of hospital_beds_total)
`oxygen_beds_occupied_r7mean`- occupied beds with oxygen supply
`mv_beds_total_r7mean`- all beds with artificial lung ventilation (DPV) (part of hospital_beds_total)
`mv_beds_occupied_r7mean`- busy beds with DPV options
`icu_beds_total_r7mean`- resuscitation and intensive care (RITS) beds (not part of hospital_beds_total!)
`icu_beds_occupied_r7mean`- busy RITS beds (not part of hospital_beds_occupied!)
`icu_ecmo_systems_total_r7mean`- total number of RITS beds with ECMO systems
`icu_ecmo_systems_in_use_r7mean `- number of occupied RITS beds with ECMO systems
`cv19_hospital_beds_assigned_r7mean`- Number of beds for COVID patients
`cv19_hospital_beds_occupied_r7mean `- number of occupied COVID beds
`cv19_oxygen_beds_assigned_r7mean`- COVID oxygen beds for patients
`cv19_oxygen_beds_occupied_r7mean`- occupied oxygen beds for COVID patients
`cv19_mv_beds_assigned_r7mean `- COVID Patient Beds with Artificial Lung Ventilation (DPV)
`cv19_mv_beds_occupied_r7mean`- occupied COVID patient beds with DPV capabilities
`cv19_icu_beds_assigned_r7mean`- COVID Patient Resuscitation and Intensive Care (RITS) Beds
`cv19_icu_beds_occupied_r7mean`- busy RITS beds for COVID patients
`cv19_icu_mv_beds_assigned_r7mean`- COVID Patient RITS Beds with Artificial Lung Ventilation (DPV)
`cv19_icu_mv_beds_occupied_r7mean`- occupied RITS beds for COVID patients with DPV capability
`cv19_icu_highflow_systems_total_r7mean`- COITS for RV beds with high-flow system
`cv19_icu_highflow_systems_in_use_r7mean`- busy COITS RITS beds with high-flow system
`cv19_icu_ecmo_assigned_r7mean`- RITS beds for COVID patients with EKMO system
`cv19_icu_ecmo_in_use_r7mean`- busy RITS beds for COVID patients with EKMO system
`cv19_admissions_r7mean`- new COVID patients (daily)
`cv19_admissions_oxygen_r7mean`- new COVID patients receiving oxygen
`cv19_admissions_mv_r7mean`- new COVID patients with artificial lung ventilation
`cv19_discharges_r7mean`- COVID patient checkouts / departures
`remdesivir_vials_used_r7mean`- used Remdesivir vials
`remdesivir_patients_total_r7mean`- Number of patients receiving Remdesivir
`remdesivir_patients_new_r7mean`- Number of new patients receiving Remdesivir